---
title: "01-data-load"
format: html
---

## Overview

This script loads and tidies the raw data exported from Gorilla for the
webcam-based Visual World Paradigm (VWP) study. There are three data
sources:

1. **Gaze data** — per-frame webcam eye-tracking samples (one CSV per
   participant/session, uploaded as a "screen" recording), merged into a
   single tidy data frame with `webgazeR::merge_webcam_files()`.
2. **Behavioral data** — trial-level accuracy, reaction time, and display
   metadata exported as one combined task CSV.
3. **Questionnaire data** — one-time, per-participant demographics
   (age, sex, dialect, race/ethnicity, vision correction, etc.)
   exported as its own CSV.

The output of this script (`edat`, `emstr`, `audio_rt`, `quest`,
`trial_data`) feeds the downstream cleaning/analysis scripts.

## Load Packages

```{r}
options(stringsAsFactors = FALSE) # never auto-convert strings to factors on import
options("scipen" = 100, "digits" = 10) # print RTs/timestamps in full instead of scientific notation

# webgazeR is not on CRAN - install once from GitHub if needed:
# install.packages("remotes")
# remotes::install_github("jgeller112/webgazeR")

library(webgazeR) # merges/processes raw Gorilla webcam gaze files
library(dplyr) # data wrangling
library(tidyr) # reshaping helpers
library(ggplot2) # plotting (used in later scripts)
library(knitr) # kable() for nicer table output in this document
```

## Load and Merge Webcam Gaze Data

### Find the raw per-participant gaze files

Gorilla exports one gaze CSV per uploaded webcam recording, plus a
separate calibration CSV per session that we don't want to analyze.

```{r}
# All CSVs exported for this task
vwp_files <- list.files(
  here::here("Data", "data_exp_272183-vall", "uploads"),
  pattern = "\\.csv$",
  full.names = TRUE
)

# Drop the calibration recordings - we only want the trial gaze data
vwp_paths_filtered <- vwp_files[!grepl("calibration", vwp_files)]

kable(head(basename(vwp_files)), col.names = "file")
```

### Merge the per-participant files into one tidy data frame

`col_map` tells `merge_webcam_files()` which raw Gorilla column names
correspond to subject ID, trial number, timestamp, and the normalized
(0-1) gaze coordinates. `screen_index = 4` selects the recording screen
that contains the VWP display (Gorilla numbers each screen shown to
participants, and the eye-tracking screen is the 4th).

```{r}
# Gorilla writes gaze files as full paths, so working directory doesn't
# strictly matter here, but merge_webcam_files() expects it to be set to
# the folder containing the uploads.
setwd(here::here("Data", "data_exp_272183-vall", "uploads"))

edat <- merge_webcam_files(
  vwp_paths_filtered,
  screen_index = 4,
  kind = "gorilla",
  col_map = list(
    subject = "`Participant Private ID`",
    trial = "`Spreadsheet Index`",
    time = "Elapsed",
    x = "`Predicted Gaze X Normalised",
    y = "Predicted Gaze Y Normalised"
  )
)
```

```{r}
kable(head(edat))
```

## Behavioral (Trial) Data

### Read the raw task-level export

This is the combined "tasks" CSV Gorilla produces, containing one row
per logged event (display change, response, audio playback, etc.) across
all participants.

```{r}
agg_eye_data <- read.csv(
  here::here("Data", "data_exp_272183-vall", "data_exp_272183-vall_tasks.csv")
)
```

### Clean, filter, and rename the trial-level fields

We keep only the rows logged during the eye-tracking display (`ET`) where
the event was an actual participant response, then rename Gorilla's
verbose column names to short, analysis-friendly labels.

```{r}
emstr <- agg_eye_data %>%

  janitor::clean_names() %>%

  # Keep only the columns needed for accuracy/RT analysis, plus
  # object_name (which quadrant image the response/fixation refers to) and
  # participant_public_id (encodes condition - see mutate() below)
  dplyr::select(
    participant_private_id,
    participant_public_id,
    correct,
    spreadsheet_tlpic,
    spreadsheet_trpic,
    spreadsheet_blpic,
    spreadsheet_brpic,
    spreadsheet_trialtype,
    spreadsheet_targetword,
    display,
    spreadsheet_tlcode,
    spreadsheet_trcode,
    spreadsheet_blcode,
    spreadsheet_brcode,
    reaction_time,
    trial_number,
    response_type,
    object_name,
    response
  ) %>%

  # Restrict to the eye-tracking display, and to rows that record the
  # participant's actual response (drops audio/fixation-only log rows)
  dplyr::filter(display == "ET", response_type == "response") %>%

  # Rename Gorilla's verbose spreadsheet column names to short labels
  dplyr::rename(
    "TL" = "spreadsheet_tlpic", # top-left image
    "TR" = "spreadsheet_trpic", # top-right image
    "BL" = "spreadsheet_blpic", # bottom-left image
    "BR" = "spreadsheet_brpic", # bottom-right image
    "targ_loc" = "object_name", # which quadrant held the target
    "subject" = "participant_private_id",
    "trial" = "trial_number",
    "acc" = "correct",
    "RT" = "reaction_time"
  ) %>%

  # RT is read in as character by default; subject/trial are IDs, not
  # continuous numbers, so treat them as factors
  mutate(
    RT = as.numeric(RT),
    subject = as.factor(subject),
    trial = as.factor(trial),
    # participant_public_id encodes condition as a leading counterbalance
    # number followed by a condition code, e.g. "1HQC" -> "HQC"; strip the
    # number to recover the condition label
    condition = as.factor(sub("^[0-9]+", "", participant_public_id))
  ) %>%

  # Drop columns that are now redundant: participant_public_id's only
  # useful content (condition) has been extracted above, and display/
  # response_type are constant ("ET"/"response") now that we've filtered
  # on them, so they carry no remaining information
  dplyr::select(-participant_public_id, -display, -response_type)

kable(head(emstr))
```

## Extract Audio Onset Times

The target word is played as audio partway through each trial. We pull
out the timestamp at which playback started for each trial so gaze data
can later be time-locked to audio (rather than trial) onset.

```{r}
audio_rt <- agg_eye_data %>%

  janitor::clean_names() %>%

  # participant_public_id/condition are already carried by emstr, and
  # object_name/response are only needed to find the onset rows below -
  # keep just what's needed to identify the row and its timestamp
  dplyr::select(
    participant_private_id,
    object_name,
    trial_number,
    reaction_time,
    response
  ) %>%

  # "Audio" rows log playback events; "audio started" marks the onset
  dplyr::filter(object_name == "Audio", response == "audio started") %>%
  dplyr::distinct() %>%

  dplyr::rename(
    "subject" = "participant_private_id",
    "trial" = "trial_number",
    "RT_audio" = "reaction_time"
  ) %>%
  # Drop object_name/response - they were only needed to filter down to
  # the onset rows and are now constant ("Audio"/"audio started"), so
  # keeping them would just be redundant, unlabeled duplicates of that
  # filter plus a name collision with emstr's own "response" column
  dplyr::select(-object_name, -response) %>%
  dplyr::mutate(RT_audio = as.numeric(RT_audio))

kable(head(audio_rt))
```

## Questionnaire (Demographics) Data

Participants also completed a consent form and a demographics
questionnaire, both exported together in one "questionnaires" CSV (one
row per completed task, not one row per participant). We keep only the
demographics task rows and tidy them into a single row per subject.

```{r}
agg_quest_data <- read.csv(
  here::here("Data", "data_exp_272183-vall", "data_exp_272183-vall_questionnaires.csv")
)
```

```{r}
quest <- agg_quest_data %>%

  janitor::clean_names() %>%

  # Consent-form rows carry no demographic answers - keep only the
  # demographics task
  dplyr::filter(task_name == "Questionnaire (new)") %>%

  # Gorilla exported two questions ("dialect" and "make up") with a
  # duplicate column name; read.csv/janitor split each into a blank
  # original and a second copy that actually holds the answer
  # (response_3_object_6_response / response_10_object_2_make_up_1) -
  # keep only the populated one and give everything readable names
  dplyr::select(
    participant_private_id,
    participant_public_id,
    age = response_1_object_4_value,
    sex = response_2_object_5_response,
    dialect = response_3_object_6_response,
    ethnicity = response_4_object_7_response,
    medication_impairment = response_9_object_3_response,
    high_prescription = response_8_object_9_response,
    english_before_4 = multiple_choice_object_10_response,
    eye_glasses = response_10_object_2_eye_glasses,
    contacts = response_10_object_2_contacts,
    make_up = response_10_object_2_make_up_1,
    hat = response_10_object_2_hat,
    eyelashes = response_10_object_2_strip_individual_eyelashes,
    race_american_indian_alaska_native = response_5_object_8_american_indian_alaska_native,
    race_asian = response_5_object_8_asian,
    race_native_hawaiian_pacific_islander = response_5_object_8_native_hawaiian_or_other_pacific_islander,
    race_black_african_american = response_5_object_8_black_or_african_american,
    race_white = response_5_object_8_white,
    race_more_than_one = response_5_object_8_more_than_one_race,
    race_decline_to_state = response_5_object_8_decline_to_state
  ) %>%

  dplyr::rename("subject" = "participant_private_id") %>%

  mutate(
    subject = as.factor(subject),
    age = as.numeric(age),
    # same condition code embedded in participant_public_id as in emstr
    condition = as.factor(sub("^[0-9]+", "", participant_public_id))
  ) %>%

  # Drop the raw public ID now that condition has been extracted from it
  dplyr::select(-participant_public_id)

kable(head(quest))
```

### Summary Table of Questionnaire Responses

For a quick demographic snapshot, we reshape `quest` to one row per
question/response pair and report how many participants gave each
response (and what percent of respondents that is). `age` is continuous
rather than categorical, so it's summarized separately below instead of
folded into the percentage table; `subject`/`condition` are IDs/group
assignment rather than questionnaire responses, so they're excluded too.

```{r}
# Helper to turn "race_more_than_one" into "Race More Than One"
title_case <- function(x) {
  words <- strsplit(gsub("_", " ", x), " ")
  vapply(words, function(w) {
    paste0(toupper(substring(w, 1, 1)), substring(w, 2), collapse = " ")
  }, character(1))
}

demo_summary <- quest %>%

  dplyr::select(-subject, -age, -condition) %>%

  # Blank strings are unanswered items; the vision/race items are coded
  # 0/1 checkboxes, so relabel them as No/Yes for a readable table
  dplyr::mutate(dplyr::across(dplyr::everything(), ~ dplyr::na_if(as.character(.x), ""))) %>%
  dplyr::mutate(dplyr::across(
    dplyr::everything(),
    ~ ifelse(.x == "1", "Yes", ifelse(.x == "0", "No", .x))
  )) %>%

  tidyr::pivot_longer(dplyr::everything(), names_to = "question", values_to = "response") %>%
  tidyr::drop_na(response) %>%

  # Count how often each response was given, as a percent of participants
  # who answered that particular question
  dplyr::count(question, response, name = "n") %>%
  dplyr::group_by(question) %>%
  dplyr::mutate(percent = round(100 * n / sum(n), 1)) %>%
  dplyr::ungroup() %>%

  dplyr::mutate(question = title_case(question)) %>%
  dplyr::arrange(question, dplyr::desc(percent))

kable(
  demo_summary,
  col.names = c("Question", "Response", "N", "Percent (%)"),
  align = c("l", "l", "r", "r")
)
```

```{r}
# Age is continuous, so report it as a single summary row instead
kable(
  data.frame(
    Question = "Age",
    Mean = round(mean(quest$age, na.rm = TRUE), 1),
    SD = round(sd(quest$age, na.rm = TRUE), 1),
    Min = min(quest$age, na.rm = TRUE),
    Max = max(quest$age, na.rm = TRUE)
  )
)
```

```{r}

trial_data_rt <- merge(emstr, audio_rt, by = c("subject", "trial"))

head(trial_data_rt) %>%
  head() %>%
  kable()
```



```{r}

samp_rate <- analyze_sampling_rate(edat, summary_stat = "Median")
```

```{r}

samp_rate$subject <- as.factor(samp_rate$subject)
samp_rate$trial <- as.factor(samp_rate$trial)


trial_data <- left_join(trial_data_rt, samp_rate, by = c("subject", "trial"))

head(trial_data)
```
